Helping Humans Align Efficiently

### 2025-09-30 doc/aligments2025/doc.tex
minor edits

\paragraph{snify}
The snify tool~\cite{KohSch:safc25} implements
a a workflow similar to the one shown in \Cref{fig:workflow}
for semantic annotation.
Both the harvester and the frontend act on \LaTeX\ documents
that use the \texttt{stex} package for semantic markup.
Using the snify tool has significantly increased
the efficiency of annotators, especially for novices who
are not deeply familiar with the domain model (i.e.\ who
do not know the identifiers by heart).

@@

Aside from verbalizations,
WikiData also has formula notations
for many concepts, including \LaTeX\ notations like \lstinline.\mathbb{N}.
(in the concept glossary, we could treat them as verbalizations in another language).
This allows us to use the same workflow to align formula fragments,
but there are significant limitations:
operators can be aligned, but their arguments are not marked, and
some operators, like the invisible multiplication in ``$ax$'',
are not represented by a glyph at all.
In sTeX, the solution is essentially to write down
a semantic representation of the formula and generate the presentation from it.
A grammar-based approach to generate the semantic representation
for ``conventional'' \LaTeX\ formulae
in collaboration with a human annotator was presented in~\cite{VreWelKam:tsmmdui24}.
For alignments, this approach could be considered too invasive.

@@

In summary, we have described a workflow for efficient human alignment
based on an concept glossary
and presented a prototypical implementation based on the snify tool
that explores this workflow
for the alignment of text and formulae in \LaTeX\ and HTML documents
with WikiData concepts.
% 
% In summary, we have demonstrated that the snify approach for sTeX annotation
% can generalize to alignments with other concept databases
% and document formats.
While the formula alignment is rather limited,
it was relatively easy to implement and could be improved in the future.
On a higher level, this shows how relatively simple and naive tools
can enable much more efficient alignment workflows,
and we hope to inspire the community to think about other tools
that could help alignment authors.
% -- be it text-to-database or database-to-database alignment.

>>>>

\paragraph{Building on snify}
The snify tool~\cite{KohSch:safc25} implements
a a workflow similar to the one shown in \Cref{fig:workflow}
for semantic annotation.
Both the harvester and the frontend act on \LaTeX\ documents
that use the \texttt{stex} package for semantic markup.
Using the snify tool has significantly increased
the efficiency of annotators, especially for novices who
are not deeply familiar with the domain model (i.e.\ who
do not know the identifiers by heart).

@@

\paragraph{Supporting Formulae and HTML}
Aside from verbalizations,
WikiData also has formula notations
for many concepts, including \LaTeX\ notations like \lstinline.\mathbb{N}.
(in the concept glossary, we could treat them as verbalizations in another language).
This allows us to use the same workflow to align formula fragments,
but there are significant limitations:
operators can be aligned, but their arguments are not marked, and
some operators, like the invisible multiplication in ``$ax$'',
are not represented by a glyph at all.
In sTeX, the solution is essentially to write down
a semantic representation of the formula and generate the presentation from it.
A grammar-based approach to generate the semantic representation
for ``conventional'' \LaTeX\ formulae
in collaboration with a human annotator was presented in~\cite{VreWelKam:tsmmdui24}.
For alignments, this approach could be considered too invasive.

@@

\paragraph{Conclusion}
In summary, we have described a workflow for efficient human alignment
based on an concept glossary
and presented a prototypical implementation based on the snify tool
that explores this workflow
for the alignment of text and formulae in \LaTeX\ and HTML documents
with WikiData concepts.
% 
% In summary, we have demonstrated that the snify approach for sTeX annotation
% can generalize to alignments with other concept databases
% and document formats.
While the formula alignment is rather limited,
it was relatively easy to implement and could be improved in the future.
On a higher level, this shows how relatively simple and naive tools
can enable much more efficient alignment workflows,
and we hope to inspire the community to think about other tools
that could help alignment authors.
% -- be it text-to-database or database-to-database alignment.
